---
title: Analysis of animals that occur in groups
author: Centre for Research into Ecological and Environmental Modelling <br> **University of St Andrews**
date-modified: "January 2024"
date-format: "MMMM YYYY"
bibliography: ../references.bib
csl: ../../style/apa.csl
---

```{r}
#| include = FALSE,
#| message = FALSE
knitr::opts_chunk$set(echo = TRUE)
solution <- FALSE
```

# Analysis of animals that occur in groups

## Description of survey

The data for this practical is simulated (so we can compare estimates derived from our analysis with truth), but the simulation is based upon a series of surveys conducted in the eastern Atlantic 2010-2019 by the U.S. National Marine Fisheries Service for a programme called Atlantic Marine Assessment Program for Protected Species [@AtlanticMarineAssessment2017].  Estimates of abundance of Risso's dolphins *(Grampus griseus)* were derived from these survey data and reported in @roberts2022.



# Read in the data
```{r}
#| label: getdata
risso <- read.csv("Risso_survey.csv")
head(risso, n=3)
```

## Data organisation


## Focus on group size

Risso's dolphins (and many other species), travel in aggregations. Explore the distribution of observed group sizes in this data set.

```{r}
#| label: groupdist
aveobs.size <- round(mean(risso$size),2)
histlabel <- paste0("Observed group sizes of Risso's dolphins\n",
                    "Mean observed size= ", aveobs.size)
hist(risso$size, nc=15, main=histlabel, xlab="Observed group size")
```

It is important to note that the mean size of detected groups is `r aveobs.size`, because we know the mean size of groups in the population was 6.

We can explore the possible reason for this difference between mean size in the population and mean size in our sample by creating this diagnostic plot. This plot should be a standard part of any analysis that involves detection of animals in groups.

```{r}
#| label: fig-moreEDA
#| fig-cap: "Diagnostic plot for existance of size bias in group size estimates."
with(risso,scatter.smooth(distance, size, pch=20,
                          lwd=2, xlab="Detection distance",
                          ylab="Size of detected group",
                          main="Risso's survey\nDiagnostic plot"))
```

Note in @fig-moreEDA the absence of small groups at distances beyond ~2nm. Beyond ~4nm, not groups are detected that are smaller than 6. This pattern is indicative of bias in the estimated size of the average group. We will see the consequence of this bias in average group size when we perform a naive analysis.

# Analysis of survey data
## Naive analysis

If we apply distance sampling to the perpendicular distances recorded to the centre of the detected groups, we will estimate the abundance of groups ($\widehat{N_s}$), corrected for imperfect detectability. To convert abundance of groups ($\widehat{N_s}$) to abundance of individuals ($\widehat{N}$), we multiply:

$$\widehat{N_s} \times \bar{s} = \widehat{N}$$
where \bar{s} is the average size of groups in the population [@Buckland2015, Section 6.3.1.3]. We do not know the average size of groups in the population, but rather we estimate it by using the average size of our detected groups.

Because there exists a field named `size` in the `risso` data frame, the `ds` software knows observations were of groups. Output from `ds` will create estimates both of $\widehat{N_s}$ and $\widehat{N}$ (if study region `Area` is also provided), and companion estimates $\widehat{D_s}$ and $\widehat{D}$.

Fit the three key function detection models to the data in the usual manner and perform model selection to choose a most appropriate model (also perform absolute goodness of fit, courtesy of `summarize_ds_models).

```{r}
#| label: naivemodels
#| message: false
#| eval: !expr solution
library(Distance)
naive.uniform <- ds(data=risso, key="unif", adjustment="cos")
naive.hn <- ds(data=risso, key="hn", adjustment = "cos") 
naive.hr <- ds(data=risso, key="hr", adjustment = NULL) # no adjustments for simplicity
knitr::kable(summarize_ds_models(naive.uniform, naive.hn, naive.hr, output="plain"),
             digits=3, caption="Model selection for models not considering size bias.")
```

All three models fit. It is a close AIC contest between the unadjusted hazard rate and the half normal with one adjustment. For our purposes, let's focus upon the hazard rate model, although the inference will be virtually identical were we to use the half normal model for our inference.

```{r}
#| label: naivehrplot
#| eval: !expr solution
plot(naive.hr, nc=40, main="Hazard rate model with no adjustments")
```

A brief look at the data summary coming from the fitted model. Evaluate the number of detections and numbers of replicate transects. 

```{r}
#| label: datasummary
#| eval: !expr solution
knitr::kable(naive.hr$dht$clusters$summary)
```

Examine the estimates of abundance of clusters ($\widehat{N_s}$) from the hazard rate model. Note how the following code carefully extracts estimates only for clusters. We could do the same for the estimated density of clusters, but omit that here.

```{r}
#| label: estimatens
#| eval: !expr solution
knitr::kable(naive.hr$dht$clusters$N)
```

Compare this estimate against the number of clusters (4333) in the population we simulated.

How well did this model estimate the number of individuals ($\widehat{N}$) in the population?

```{r}
#| label: estimaten
#| eval: !expr solution
knitr::kable(naive.hr$dht$individuals$N)
```

Compare this estimate against the number of individuals in the population (26000). The reason for this lies in the estimation of average group size in the population, also estimated in the model object by the average size of detected groups:

```{r}
#| label: estimatesbar
#| eval: !expr solution
knitr::kable(naive.hr$dht$Expected.S)
```

Compare this estimate against  the estimate we produced during our exploratory data analysis plotting the histogram of sizes of observed groups: `r aveobs.size`.

## Analysis adjusting for size bias problem

Because we recognize that the size of clusters influences probability of their inclusion in our sample, we can incorporate this concept in the detection function model we construct. As introduced in [Module 8](../../08-covariates/covariateslanding.qmd), we can add covariates in addition to perpendicular distance into our detection function models. This is what we will do to perhaps overcome the size bias problem.

:::{.callout-note collapse=false appearance='default' icon=true}
## Aside: average group size as derived parameter

- When using group size as a covariate we no longer estimate average group size directly from observed group sizes. Instead, the `ds` function uses something called Horwitz-Thompson-like estimators to first estimate the abundance of groups. Then, using the same estimation approach, `ds` estimates the abundance of individuals. 
- After the two estimates $\widehat{N_s}$ and $\widehat{N}$ are produced, their ratio is computed. This ratio is a less biased estimate of average group size in the population.
- See the next-to-last slide in [Module 8 lecture](../../08-covariates/Mod8-2023-covariates.pdf)
:::

```{r}
#| label: clevermodels
#| eval: !expr solution
clever.hn <- ds(data=risso, key="hn", formula = ~size) 
clever.hr <- ds(data=risso, key="hr", formula = ~size)
```

Contrast AIC scores between hazard rate and half normal models with and without the `size` covariate.

```{r}
#| label: AICscores
#| eval: !expr solution
AIC(naive.hn, naive.hr, clever.hn, clever.hr)
```

Compare relative and absolute fit of hazard rate and half normal key functions that include the `size` covariate.

```{r}
#| label: sumtable
#| eval: !expr solution
knitr::kable(summarize_ds_models(clever.hn, clever.hr, output="plain"), digits=3,
             caption = "Comparison of hazard rate and half normal models incorporating group size as covariate.")
```

Using the best model based upon AIC, repeat the model output interrogation you conducted (above) with the naive models: examine the estimates of $\widehat{N_s}$, $\widehat{N}$ and $E(s)$ (defined as the expected value of group size in the population):

```{r}
#| label: cleverns
#| eval: !expr solution
knitr::kable(clever.hr$dht$clusters$N)
```

Compare this estimate against the number of clusters (4333) in the population we simulated.

How well did this model estimate the number of individuals ($\widehat{N}$) in the population?

```{r}
#| label: clevern
#| eval: !expr solution
knitr::kable(clever.hr$dht$individuals$N)
```

Compare this estimate against the number of individuals in the population (26000). The reason for this lies in the estimation of average group size in the population, also estimated in the model object by the average size of detected groups:

```{r}
#| label: cleversbar
#| eval: !expr solution
knitr::kable(clever.hr$dht$Expected.S)
```

